Rao-Blackwellized smoother for Conditionally Linear and Gaussian models

### 2016-11-15 TexArxiv/clgm.tex
MAJ

%where (setting $\Delta t_i = t_i-t_{i-1}=1$):

@@

where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\Delta = t_i-t_{i-1}$):

@@

\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta + \alpha_{a_{i-1}}[1-\rme^{-\kappa \Delta}]/\kappa\\
\alpha_{a_{i-1}} [1-\rme^{-\kappa \Delta}] \end{pmatrix}\eqsp,\\

@@

\begin{pmatrix} 1 & -[1-\rme^{-\kappa \Delta}]/\kappa \\ 0 & \rme^{-\kappa \Delta}  \end{pmatrix} \eqsp,\\
\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta + \eta^2_{a_{i-1}}\left\{\Delta  + (1-\rme^{-2\kappa \Delta})/(2\kappa) - 2(1-\rme^{-\kappa \Delta})/\kappa\right\}/\kappa^2 \\
&\hspace{4.4cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa \Delta})/\kappa \right\}/\kappa\eqsp, \\
\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa \right)\left(1-\rme^{-\kappa \Delta}} \right)/\kappa} \\
&\hspace{7.8cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa \Delta}} \right)/(2\kappa^2)\eqsp,\\
\barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp,\\
\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa \Delta} \right)/(2\kappa)\eqsp.

@@

%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \Delta t_i}]/\kappa\\
%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \Delta t_i}] \end{pmatrix}\eqsp,

@@

The observations are Wednesday future contracts of the West Texas Intermediate crude oil (WTI) traded in the Chicago Mercantile Exchange (CME) from $11$ January 1995 to $13$ November 2013. The contracts are numbered $F_1, F_2, \ldots, F_{36}$ where $F_1$ (or front month) is the earliest delivery future contract, $F_2$ is the second earliest delivery future contract and so on. Among these $36$ contracts, the four future contracts: $F_1, F_4, F_6, F_{13}$ are used since their trading volumes and their impacts on the Term Structures are the most important ($F_1$ is the most liquid contract, $F_{13}$ characterizes the gap between prices over a one year period, $F_4$ and $F_6$ are intermediate future contracts that are mostly traded). As in \cite{almansour:2016}, we consider that each  future contract has a fixed time to maturity: $F_1, F_4, F_6, F_{13}$ have time to  maturity $4$, $16$, $26$, $56$ weeks.  Our time series contains $n =975$ weekly data with $534$ in backwardation and $441$ in contango (the backwardation effect is more frequent with crude oil data). At each time $t_i = i\Delta t$, with $\Delta t = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\Delta t,m_1}), \ldots, \ln(F^{(market)}_{i\Delta t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.

@@

where $\mathsf{B}_0	= \begin{pmatrix}1 & 0 \end{pmatrix}$ and  $\mathsf{B}_m    = \mathsf{B}_{m-1} T$ so that  $ \mathsf{B}_m= \begin{pmatrix} 1 & -\left(1-\rme^{-\kappa m \Delta t}\right)/\kappa \end{pmatrix}$ and
%$B_\tau$ is the $1 \times 2$ matrix given by $B_{\tau} \eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}$ and
%for all $1\le j\le J$, $A_0(j) \eqdef 0$. Define $F_{i,m} \eqdef F_{i\Delta t, m \Delta t }$ and $\mathcal{F}_i$ the $\sigma$-field generated by $(a_k)_{0\le k \le i}$ and $(Z_k)_{0\le k \le i}$. By definition of future prices and the assumption that there is no regime switching between $t_i$ and $t_{i+1}$, we have:
%%\begin{align*}
%%A_{t,\tau} &\eqdef r \tau - \int^{\tau}_0 \alpha_{a_{t+s}} \left(1- e^{-\kappa(\tau-s)} \right)\rmd s + (1/\kappa^2)\int^{\tau}_0 \eta^2_{a_{t+s}} \left(1-e^{-\kappa (\tau-s)}\right)^2 \rmd s \\
%%&\hspace{7cm}  - (2/\kappa)\int^{\tau}_0 \rho_{a_{t+s}} \eta_{a_{t+s}}\sigma_{a_{t+s}} \left( 1 -e^{-\kappa \left(\tau-s\right)} \right)\rmd s \eqsp,\\
%%B_{\tau} &\eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}\eqsp.
%%\end{align*}
%F_{i,m} &= \exp (A_m(a_i) + B_m Z_i)\eqsp,\\
%&=\mathbb{E}\left[\exp (A_{m-1}(a_{i+1}) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right]\eqsp,\\
%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp (A_{m-1}(k) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right] \eqsp,\\
%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp \left(A_{m-1}(k) + B_{m-1} ( d_{a_i} + T Z_i + H_{a_i} \epsilon_{i+1} ) \right)\middle|\mathcal{F}_i\right] \eqsp,\\
%&= \sum^J_{k=1} q(a_i,k) \exp \left(A_{m-1}(k) + B_{m-1} d_{a_i} + B_{m-1} \barH_{a_i} B'_{m-1}/2 + B_{m-1} T Z_i \right)\eqsp.
%From the above equality, we obtain,

@@

\begin{enumerate}%[label=\roman*)]

@@

The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi(1) = \pi(2) = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime. The other parameters are initialized as shown in the following table:
\begin{figure}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
		$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
		\hline
		5.0 & 0.1 & -0.05 & 0.4 & 0.4 & 0.5 & 0.5 & 0.75 & 0.65 & 0.0 & 0.0 & 0.0 & 0.1 & 0.1 & 0.1 & 0.1
	\end{tabular}
	\caption{Initial values for the EM algorithm.}
\end{figure}\\
The number of particles is set to $N = 100$, $\Delta t = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
\[
\mu_1 = \begin{pmatrix} \ln F^{(market)}_{1,4} & r-\cfrac{\ln F^{(market)}_{1,16}-\ln F^{(market)}_{1,4}}{(16-4)\Delta t} \end{pmatrix}\quad \mbox{and}\quad \Sigma_1 = \begin{pmatrix} 0.05 & 0 \\ 0 & 0.05 \end{pmatrix}\eqsp.
\]
The CMA-ES algorithm is used with an initial standard deviation for the parameters $\sigma_{cmaes} = 0.005$, a number of selected search points $\mu_{cmaes} = 20$ and a population size $\lambda_{cmaes} = 100$. The algorithm is stopped after $10000$ iterations.
%and its bounded conditions are \\
%\begin{tabular}{cccccccccccccc}
%PAR & $\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
%\hline
%INF & 0.01 & -10.0 & -10.0 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 1e-10 & 1e-10 & 1e-10 & 1e-10 \\
%SUP & 1e10 & 10.0 & 10.0 & 1e10 & 1e10 & 1e10 & 1e10 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 \\
%\end{tabular}
In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters. For the CME WTI Crude Oil, backwardation effect is more frequent than contango effect so that $\alpha_1$ should be greater than $\alpha_2$. Therefore, this condition is imposed for all simulations in the CMA-ES algorithm. The results after $2500$ iterations of the EM algorithm are given in Figure~\ref{fig:CMEfinalestimates}.
\begin{table}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c||c|clc|c|c||}
		Parameter&$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
		\hline
		Value&2.6378 & 0.0889 & -0.0281 & 0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
		\hline
		Std. Dev& 0.1999 & 0.004248 & 0.00149 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052   \\
	\end{tabular}
	\vspace{1.cm}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|||}
		Parameter&$\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
		\hline
		Value&0.1712 & 0.2422	& -0.1731 & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
		\hline
		Std. Dev& 0.004147 & 0.03243 & 0.01973 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4
	\end{tabular}
	\caption{Final estimates after 3500 iterations.}
	\label{fig:CMEfinalestimates}
\end{table}

>>>>

%where (setting $\tau t_i = t_i-t_{i-1}=1$):

@@

where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\tau = t_i-t_{i-1}$):

@@

\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \tau + \alpha_{a_{i-1}}[1-\rme^{-\kappa \tau}]/\kappa\\
\alpha_{a_{i-1}} [1-\rme^{-\kappa \tau}] \end{pmatrix}\eqsp,\\

@@

\begin{pmatrix} 1 & -[1-\rme^{-\kappa \tau}]/\kappa \\ 0 & \rme^{-\kappa \tau}  \end{pmatrix} \eqsp,\\
\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \tau + \eta^2_{a_{i-1}}\left\{\tau  + (1-\rme^{-2\kappa \tau})/(2\kappa) - 2(1-\rme^{-\kappa \tau})/\kappa\right\}/\kappa^2 \\
&\hspace{4.4cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \tau t_i - (1-\rme^{-\kappa \tau})/\kappa \right\}/\kappa\eqsp, \\
\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa \right)\left(1-\rme^{-\kappa \tau} \right)/\kappa + \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa \tau} \right)/(2\kappa^2)\eqsp,\\
\barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp, \quad \barH_{a_{i-1}}(2,2) = \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa \tau} \right)/(2\kappa)\eqsp.

@@

%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \tau t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \tau t_i}]/\kappa\\
%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \tau t_i}] \end{pmatrix}\eqsp,

@@

The observations are Wednesday future contracts of the West Texas Intermediate crude oil (WTI) traded in the Chicago Mercantile Exchange (CME) from $11$ January 1995 to $13$ November 2013. The contracts are numbered $F_1, F_2, \ldots, F_{36}$ where $F_1$ (or front month) is the earliest delivery future contract, $F_2$ is the second earliest delivery future contract and so on. Among these $36$ contracts, the four future contracts: $F_1, F_4, F_6, F_{13}$ are used since their trading volumes and their impacts on the Term Structures are the most important ($F_1$ is the most liquid contract, $F_{13}$ characterizes the gap between prices over a one year period, $F_4$ and $F_6$ are intermediate future contracts that are mostly traded). As in \cite{almansour:2016}, we consider that each  future contract has a fixed time to maturity: $F_1, F_4, F_6, F_{13}$ have time to  maturity $4$, $16$, $26$, $56$ weeks.  Our time series contains $n =975$ weekly data with $534$ in backwardation and $441$ in contango (the backwardation effect is more frequent with crude oil data). At each time $t_i = i\tau$, with $\tau = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\tau t,m_1}), \ldots, \ln(F^{(market)}_{i\tau t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.

@@

where $\mathsf{B}_0	= \begin{pmatrix}1 & 0 \end{pmatrix}$ and  $\mathsf{B}_m    = \mathsf{B}_{m-1} T$ so that  $ \mathsf{B}_m= \begin{pmatrix} 1 & -\left(1-\rme^{-\kappa m \tau t}\right)/\kappa \end{pmatrix}$ and

@@

\begin{enumerate}

@@

The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi(1) = \pi(2) = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime. The other parameters are initialized as shown in the following table:
\begin{figure}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
		$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
		\hline
		5.0 & 0.1 & -0.05 & 0.4 & 0.4 & 0.5 & 0.5 & 0.75 & 0.65 & 0.0 & 0.0 & 0.0 & 0.1 & 0.1 & 0.1 & 0.1
	\end{tabular}
	\caption{Initial values for the EM algorithm.}
\end{figure}\\
The number of particles is set to $N = 100$, $\tau  = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
\[
\mu_1 = \begin{pmatrix} \ln F^{(market)}_{1,4} & r-\cfrac{\ln F^{(market)}_{1,16}-\ln F^{(market)}_{1,4}}{(16-4)\tau} \end{pmatrix}\quad \mbox{and}\quad \Sigma_1 = \begin{pmatrix} 0.05 & 0 \\ 0 & 0.05 \end{pmatrix}\eqsp.
\]
The CMA-ES algorithm is used with an initial standard deviation for the parameters $\sigma_{cmaes} = 0.005$, a number of selected search points $\mu_{cmaes} = 20$ and a population size $\lambda_{cmaes} = 100$. The algorithm is stopped after $10000$ iterations.
%and its bounded conditions are \\
%\begin{tabular}{cccccccccccccc}
%PAR & $\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
%\hline
%INF & 0.01 & -10.0 & -10.0 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 1e-10 & 1e-10 & 1e-10 & 1e-10 \\
%SUP & 1e10 & 10.0 & 10.0 & 1e10 & 1e10 & 1e10 & 1e10 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 \\
%\end{tabular}
In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters. For the CME WTI Crude Oil, backwardation effect is more frequent than contango effect so that $\alpha_1$ should be greater than $\alpha_2$. Therefore, this condition is imposed for all simulations in the CMA-ES algorithm. The results after $2500$ iterations of the EM algorithm are given in Figure~\ref{fig:CMEfinalestimates}.
\begin{table}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c||c|clc|c|c||}
		Parameter&$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
		\hline
		Value&2.6378 & 0.0889 & -0.0281 & 0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
		\hline
		Std. Dev& 0.1999 & 0.004248 & 0.00149 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052   \\
	\end{tabular}
	\vspace{1.cm}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|||}
		Parameter&$\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
		\hline
		Value&0.1712 & 0.2422	& -0.1731 & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
		\hline
		Std. Dev& 0.004147 & 0.03243 & 0.01973 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4
	\end{tabular}
	\caption{Final estimates after 3500 iterations.}
	\label{fig:CMEfinalestimates}
\end{table}
